---
title: Multivariate Normal CDFs and Their Derivatives
author: Benjamin Christoffersen
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  md_document:
    variant: markdown_github
    pandoc_args: --webtex=https://latex.codecogs.com/svg.latex?
bibliography: ../README.bib
---

# Multivariate Normal CDFs and Their Derivatives

Setup.

```{r setup}
set.seed(1L)
options(digits = 4)

# sample data
get_sim <- function(){
  within(list(), {
    p <- 4L
    sigma <- drop(rWishart(1, 2L * p, diag(p)))
    mu <- drop(crossprod(chol(sigma), rnorm(p)))
  })
}

invisible(list2env(get_sim(), envir = environment()))
```

```{r def_R_func}
#####
# define function to approximate rectangular multivariate normal CDFs
# where the lower bound are -Inf and upper bounds are zero
my_pmvnorm <- function(mean, sigma, nsim, h = function(...) 1){
  #####
  # setup
  n <- length(mean)
  sigma_chol <- chol(sigma)
  draw_trunc_mean <- function(b, u)
    qnorm(pnorm(b) * u)
  nm1 <- n - 1L

  #####
  # simulate
  sims <- replicate(nsim, {
    u <- runif(n)
    draw <- numeric(n)
    out <- 1.

    for(i in 1:n){
      b <- if(i == 1L)
        -mean[i]
      else
        -mean[i] - drop(sigma_chol[1:(i - 1L), i] %*% draw[1:(i - 1L)])
      b <- b / sigma_chol[i, i]

      draw[i] <- draw_trunc_mean(b, u[i])

      out <- out * pnorm(b)
    }

    h(draw, mean, sigma_chol) * out
  }, simplify = "array")

  # return MC estimate
  if(is.vector(sims))
    return(mean(sims))
  rowMeans(sims)
}
```

Approximate CDF.

```{r aprx_cdf}
library(mvtnorm)
my_pmvnorm(mu, sigma, 10000L)
pmvnorm(upper = rep(0, p), mean = mu, sigma = sigma)
```

Approximate derivative of CDF w.r.t. the mean.

```{r arpx_deriv}
my_pmvnorm(mu, sigma, 10000L, h = function(draw, mean, sigma_chol)
  forwardsolve(sigma_chol, draw, upper.tri = TRUE))

library(numDeriv)
drop(jacobian(function(x){
  set.seed(31958429L)
  pmvnorm(upper = rep(0, p), mean = x, sigma = sigma)
}, mu))
```

Approximate derivative of log CDF w.r.t. the mean.

```{r aprx_log_deriv}
out <- my_pmvnorm(mu, sigma, 10000L, h = function(draw, mean, sigma_chol)
  c(1, forwardsolve(sigma_chol, draw, upper.tri = TRUE)))
out[-1] / out[1]

drop(jacobian(function(x){
  set.seed(31958429L)
  log(pmvnorm(upper = rep(0, p), mean = x, sigma = sigma))
}, mu))
```

## Computation Times
Make C++ example and compare computations times.

```{r cmp_cpp_func, engine = 'Rcpp'}
#include <stdexcept>
#include <array>
#include <limits>

// [[Rcpp::depends("RcppArmadillo")]]
#include <RcppArmadillo.h>

template<bool draw>
inline std::array<double, 2> draw_trunc_mean
  (double const b, const double u){
  double const qb = R::pnorm5(b, 0, 1, 1L, 0L);
  if(draw)
    return { qb, R::qnorm5(qb * u, 0, 1, 1L, 0L) };
  return { qb, std::numeric_limits<double>::quiet_NaN() };
}

// [[Rcpp::export]]
Rcpp::NumericVector my_pmvnorm_cpp(
    arma::vec const &mean_in, arma::mat const &sigma_in, 
    unsigned const nsim, double const eps){
  arma::uword const p = mean_in.n_elem;
  if(sigma_in.n_cols != p or sigma_in.n_rows != p)
    throw std::invalid_argument("invalid sigma");
  
  /* re-scale */
  arma::vec const sds = arma::sqrt(arma::diagvec(sigma_in));
  arma::vec const mean = mean_in / sds;
  
  arma::mat const sigma = ([&](){
    arma::mat out = sigma_in;
    out.each_row() /= sds.t();
    out.each_col() /= sds;
    
    return out;
  })();
  
  arma::mat const sigma_chol = arma::chol(sigma);
  
  double out(0.), di(0.), M(0.);
  arma::vec draw(p), u(p);
  unsigned i;
  unsigned const min_run(p * 25L);
  constexpr double alpha(2.576);
  for(i = 0; i < nsim;){
    unsigned const i_max = i + min_run;
    for(; i < i_max; ++i){
      auto func = [&](arma::vec const &u){
        double w(1.);
        unsigned j;
        for(j = 0; j < p - 1; ++j){
          double b(-mean[j]);
          for(unsigned k = 0; k < j; ++k)
            b -= sigma_chol.at(k, j) * draw[k];
          b /= sigma_chol.at(j, j);
          
          auto const draw_n_p = draw_trunc_mean<true>(b, u[j]);
          w       *= draw_n_p[0];
          draw[j]  = draw_n_p[1];
        }
        
        double b(-mean[j]);
        for(unsigned k = 0; k < j; ++k)
          b -= sigma_chol.at(k, j) * draw[k];
        b /= sigma_chol.at(j, j);
        
        auto const draw_n_p = draw_trunc_mean<false>(b, u[j]);
        w *= draw_n_p[0];
        
        return w;
      };
      
      u.for_each([](double &val) { return val = unif_rand(); });
      double w = func(u);
      u.for_each([](double &val) { return val = 1 - val; });
      w += func(u);
      w /= 2;
      
      /* use Welfords online algorithm */
      double const old_diff = (w - out);
      di  += 1.;
      out += old_diff / di;
      M   += old_diff * (w - out);
      
    }
    
    if(alpha * std::sqrt(M / di / di) < eps)
      break;
  }
  
  Rcpp::NumericVector ret_val(1L);
  ret_val[0L] = out;
  ret_val.attr("error") = std::sqrt(M / di / di);
  ret_val.attr("n_sim") = i;
  return ret_val;
}
```

We also use the implementation in the package. Test the C++ function.

```{r test_cpp}
# approximate CDF
library(mvtnorm)
library(mixprobit)
abs_eps <- 1e-4
n_sim <- 10000000L

my_wrap <- function()
  my_pmvnorm_cpp(mu, sigma, n_sim, abs_eps)
cm_wrap <- function()
  pmvnorm(
    upper = rep(0, p), mean = mu, sigma = sigma, 
    algorithm = GenzBretz(abseps = abs_eps, releps = -1, 
                          maxpts = n_sim))
# differs by using R's pnorm and qnorm
my_wrap_2 <- function()
  mixprobit:::my_pmvnorm_cpp(mean_in = mu, sigma_in = sigma, 
                             eps = abs_eps, nsim = n_sim)
cm_wrap_2 <- function()
  mixprobit:::pmvnorm_cpp(
    lower = rep(-Inf, p), 
    upper = rep(0, p), mean = mu, cov = sigma, 
    abseps = abs_eps, releps = -1, maxpts = n_sim)

print(my_wrap  (), digits = 6)
print(my_wrap_2(), digits = 6)
print(cm_wrap  (), digits = 6)
print(cm_wrap_2(), digits = 6)

sd(om  <- replicate(30L, my_wrap  ()))
sd(om2 <- replicate(30L, my_wrap_2()))
sd(cm  <- replicate(30L, cm_wrap()))
sd(cm2 <- replicate(30L, cm_wrap_2()$value))

cat(sprintf("%.8f\n%.8f\n%.8f\n%.8f\n%.8f\n%.8f\n%.8f\n%.8f\n", 
            mean(om), mean(om2), mean(cm), mean(cm2), 
            log(mean(om)), log(mean(om2)), log(mean(cm)), log(mean(cm2))))

# run benchmark 
set.seed(2)
bench_data <- replicate(100, {
  list2env(get_sim(), parent.env(environment()))

  sapply(list(
    `pmvnorm`        = cm_wrap, 
    `pmvnorm pkg`    = cm_wrap_2, 
    `my pmvnorm`     = my_wrap, 
    `my pmvnorm pkg` = my_wrap_2
  ), function(func){
    system.time(replicate(10L, func())) / 5L
  })
})

# show computation times
apply(bench_data["user.self", , ], 1, function(x)
  c(mean = mean(x), sd = sd(x), quantile(x, seq(0, 1, length.out = 5))))
```

Clearly, do not directly use the GHK method used by @Hajivassiliou96 or 
@Genz92 (two very similar and maybe independent suggestions). 
The main difference between `my pmvnorm` and `my pmvnorm pkg` is that the 
latter reorders the variables. 
Use what Alan 
Genz has implemented in `mvtnorm::pmvt`. See 
@Niederreiter72, @Keast73, @Cranley76 and the `MVKBRV` subroutine in 
the `mvt.f` file. Using this subroutine to approximate the derivatives 
should be fairly straight forward.

## References
